One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-08 paper/discussion.tex
updating...

  Fragmentation can be avoided if the disc is heated to maintain 
  $Q>Q_c$, the threshold for fragmentation \citep[$Q_c\simeq
  1.4$ for isothermal discs,][]{mayer04}. This may be 
  possible in the outer parts of protoplanetary discs due to
  stellar irradiation \citep{rafikov07}. Sufficiently strong
  external irradiation is expected to suppress the linear gravitational
  instabilities altogether \citep{rice11}.

  %chiang and goldreich 
  The background torque may therefore exist in the outer
  parts of protoplanetary discs that are strongly irradiated, such
  that the disc temperature is set externally with a non-zero radial
  gradient. Of course, if external irradiation sets a strictly
  isothermal outer disc, then the background torque vanishes.

@@

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the adopted disc
  structure. Thus, in our disc models the main role of disc structure
  and self-gravity is to 
  allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque density.

  In order to confine an $m=1$ spiral between two $Q$-barriers, we
  should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
  \begin{align}\label{qb_cond}
    \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
  \end{align}
  Then it may be possible to have two $Q$-barriers when the $Q$ profile
  of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
  for decreasing (increasing) $R$.

  %in fall/mass loading? 
  The surface density bump adopted in our numerical simulations might
  represent 
  `dead zone' in a protoplanetary
  disc, where the magneto-rotational 
  instability is ineffective for angular momentum transport
  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
  accretion rate. The dead zone may become 
  self-gravitating with sufficient mass built-up
  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

@@

{\bf
  \subsection{Motion of the central star}

>>>>

  Fragmentation can be avoided if the disc is heated to maintain 
  $Q>Q_c$, the threshold for fragmentation \citep[$Q_c\simeq
  1.4$ for isothermal discs,][]{mayer04}. This may be 
  possible in the outer parts of protoplanetary discs due to
  stellar irradiation \citep{rafikov09,zhu12}. Sufficiently strong
  external irradiation is expected to suppress the linear gravitational
  instabilities altogether \citep{rice11}.

  %chiang and goldreich 
  The background torque may thus exist in the outer
  parts of protoplanetary discs that are strongly irradiated, such
  that the disc temperature is set externally with a non-zero radial
  gradient. Of course, if external irradiation sets a strictly
  isothermal outer disc \citep[e.g.][]{boley09}, then the background
  torque vanishes.

@@

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the adopted surface
  density bump (Eq. \ref{sig_bump}).  
  Thus, in our disc models the main role of disc structure 
  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque.

  In order to confine an $m=1$ mode between two $Q$-barriers, we
  should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$,
  this amounts to 
  \begin{align}\label{qb_cond}
    \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
  \end{align}
  Then two $Q$-barriers may exist when the $Q$ profile
  rises more rapidly (decays more slowly) than $R^{-3/2}$
  for decreasing (increasing) $R$.

  %in fall/mass loading? 
  The surface density bump adopted in our numerical simulations might
  represent, for example, mass built-up in a 
  `dead zone' in protoplanetary 
  discs, where the magneto-rotational 
  instability is ineffective for angular momentum transport
  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
  accretion rate. The dead zone becomes 
  self-gravitating with sufficient mass built-up 
  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13}.

%  There is some attempt to do so when discussing dead zones, but I
%  doubt that the conditions there are actually suitable as such a
%  region would either be stable, or cool very slowly and be
%  susceptible to the growth of long-lived, high-m-mode spiral
%  structure.

@@

{\bf
  \subsection{Motion of the central star} 
  For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

### 2015-01-08 paper/model.tex
updating...

For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

@@

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $s$ is the power-law index describing the smooth disc and
$\Sigma_\mathrm{ref}$ is a 
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_{1,2} = \delta R \times H(R_{1,2})$.

% {\bf The bump function mimics mass built-up in $R\in[R_1,R_2]$.

% }

% In the case of protoplanetary discs, which have complex structure
% \citep{armitage10}, GI can also occur in `dead
% zones' --- sub-portions of the disc where the magneto-rotational
% instability is ineffective for angular momentum transport
% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
% accretion rate. Dead zones may then become
% self-gravitating with sufficient mass built-up
% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

%This might represent, for example, mass  
 % built-up in a circumstellar or protoplanetary
 % disc at the outer edge of a `dead zone', in which there
  %is inefficient magneto-hydrodynamic turbulence to enable accretion.

>>>>

@@

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $s$ is the power-law index describing the smooth disc and
$\Sigma_\mathrm{ref}$ is a 
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}\label{sig_bump}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_{1,2} = \delta R \times H(R_{1,2})$.
